\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-31 paper.tex
para on the true arg and relation to models

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form, the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages was formalised as a
graph by Griffiths and colleagues~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inference from a set of sampled genome sequences.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Another concept of an ARG that we must be careful to distinguish,
is that of the \emph{actual} ARG underlying a sample of genomes
and ultimately an entire species. This ARG is the true history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Just as this true, ideal ARG unquestionably exists,
it can also never be entirely known,
% NOTE: preempting some discussion here:
% Mutation rate is (a) not high enough in existing species to
% resolve, (b) probably couldn't be if organisms are to survive,
% and (c) such a high mutation rate would probably even obscure
% the more ancient past.
because of a fundamental lack of mutational information.

The scale of modern datasets makes the
relationship between the true ARG and
population genetic models important to articulate.
The coalescent with recombination (henceforth, ``coalescent'')
is a mathematical model that defines a distribution of ARGs
for a given set of parameters.
It has proven to be a useful and robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
but many modern datasets have properties that grossly
violate its assumptions.
One key assumption is that
sample size $n$ is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is an order of magnitude \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from this
assumption, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
A model assuming a single $N_e$ would be a
drastic over-simplification in these cases of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure in these
datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
Another core assumption of the coalescent model is that the genome
(or at least the region under study) is short enough that the number of extant
% TODO do we need to explain this a bit more?
lineages remains much smaller than $N_e$ at all times in the past.
High-quality whole genome assemblies are now available
for many species % Can we say something more specific here?
and projects are under way to obtain
them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth}.
Vast quantities of whole genome sequence data across a broad range
of species exists today.

The true ARG underlying a modern population-scale dataset
is therefore not a sample from the coalescent, or any other
population genetics model.

% Mathematical models are insufficient to describe the complexities
% of biological reality when sampling is at population scale,
% and the true ARG can never be known in its entirity.

A reasonable goal to set, then, is to estimate \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
by neither fully imposing the constraints of a mathematical
model or trying estimate the true ARG is its compete detail, they
have made substantial progress. However, our vocabularly
and mathematical formalisms, rooted in both the ideal and
mathematical concepts of an ARG, are insufficient to capture the
richness of these new structures.

% This ideal ARG is also not a sample from the coalescent
% or any other population genetics model: by necessity, and design,
% such mathematical constructs are caricatures of biological reality,
% allowing us to draw general conclusions about population processes
% without the burden of unnecessary detail.
% The scale of present-day datasets breaks fundamental assumptions
% made by such models, and we must therefore be careful to articulate
% our goals in inference.

% pedantically, ARG is not really a "word"
The term ``ARG'' is now frequently used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
as a data structure
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage is a useful generalisation,
but is not compatible with the available formal
data structure definitions
derived from the original Griffiths approach.
This mismatch between usage and definition has led to significant
problems, which if not addressed
may impede progress in applying ARGs to understand the biological world.

% % Para 3: what is the problem?
The classical formal definition of an ARG is derived from the
stochastic process, and expressed in terms of events:
the common ancestor and recombination events
that occur in the history of a sample. We show that this approach
to representing ARGs (which we call the ``event ARG'', or eARG
encoding) is inherently limiting, insufficient to represent
the complexities of modern datasets, and incompatible with
the outputs of modern large-scale simulation and inference
methods~\citep{kelleher2016efficient,speidel2019method,
kelleher2019inferring,zhang2023biobank}.
We address these problems by providing a more flexible formulation
of an ARG data structure, which we call the ``genome ARG''
or gARG encoding. We show that it is fully compatible with the
classical eARG approach, while providing much more flexibility
in how genetic inheritance can be represented,
and encompasses the outputs of modern methods.
Moreover we show that, in contrast with the eARG encoding which
requires precise estimates of each recombination event,
the gARG encoding allows us to systematically describe uncertainty
about the temporal ordering of multiple recombinations.

>>>>

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form, the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages was formalised as a
graph by Griffiths and colleagues~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inference from a set of sampled genome sequences.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Another concept of an ARG that we must be careful to distinguish
is that of the \emph{actual} ARG underlying a sample of genomes
and ultimately an entire species. This ARG is the true history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Just as this true, species ARG unquestionably exists,
it can also never be entirely known.
% TODO: I'm not happy with the phrasing here, anyone want to jump in?
Even if a species
had a sufficiently high mutation rate to identify every
recent branching point uniquely (and somehow managed to survive)
this high rate of mutations would inevitably obscure
more ancient structure.
% Must be a citation for this?
Similarly, it seems unlikely that the details
of ancient recombinations can ever be known with much precision.

% Popgen models and the species ARG
Species-level ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
and exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of this cohort.
UKB is just one of many such projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}, and
agricultural datasets are even more
detailed~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
While we are not sequencing the entire species yet,
we are certainly sampling at a scale where the assumptions of
population genetics models are substantially broken.
By necessity and design, such mathematical models are caricatures
of biological reality,
allowing us to draw general conclusions about population processes
without the burden of unnecessary detail.
The coalescent with recombination (henceforth, ``coalescent'')
is a powerful model that defines a distribution of ARGs
for a given set of parameters.
A key assumption of the coalescent is that
sample size is much less than the effective population size ($N_e$)
and although the model has proved surprisingly robust to violations
of this assumption~\citep{wakeley2012gene,bhaskar2014distortion,
nelson2020accounting},
sample size can now be an order of magnitude or more \emph{larger} than the
usually assumed $N_e$ values.
A model assuming a single $N_e$ would be a
drastic over-simplification in these cases of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure or the vagaries of history inherent in these
datasets.

% What is the problem?
The true ARG underlying a modern population-scale dataset
therefore cannot be known in its entirety, and the value
of attempting to describe the intricate details of the recent past
through mathematical models is questionable.
A pragmatic goal that we may set, then, is to infer \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
% TODO should we mention SARGE? I don't think it's in the same class, but
% maybe I'm wrong?
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
which estimate an approximate version of the true
ARG, informed, but not strictly constrained, by population genetic
models and parsimony.
While these methods have made substantial progress there
are, however, some significant problems in this emerging field.
Our terminology and mathematical formalisms for ARGs,
ultimately derived from the coalescent,
are ill-suited to describing these approximate structures.
% TODO add some citations of examples?
Discussions are forced to contrast a ``full'' ARG, containing
a precise description of every recombination or coalescence event,
with a sequence of disconnected local trees.
This view is depauperate, obscuring many shades of important
and interesting detail between these two extremes.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
as informally
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical description of an ARG data structure.
Classical formalisms, based on the Griffiths approach, are insufficient to
describe these rich details.
Given that ARGs have already been estimated for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
it is vital that our formal methods and terminology catch up
with this reality.

% % pedantically, ARG is not really a "word"
% The term ``ARG'' is now frequently used in a general
% sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
% schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
% as a data structure
% encompassing the varied outputs from modern simulation and
% inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
% kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
% This broad usage is a useful generalisation,
% but is not compatible with the available formal
% data structure definitions
% derived from the original Griffiths approach.
% This mismatch between usage and definition has led to significant
% problems, which if not addressed
% may impede progress in applying ARGs to understand the biological world.

% % % Para 3: what is the problem?
% The classical formal definition of an ARG is derived from the
% stochastic process, and expressed in terms of events:
% the common ancestor and recombination events
% that occur in the history of a sample. We show that this approach
% to representing ARGs (which we call the ``event ARG'', or eARG
% encoding) is inherently limiting, insufficient to represent
% the complexities of modern datasets, and incompatible with
% the outputs of modern large-scale simulation and inference
% methods~\citep{kelleher2016efficient,speidel2019method,
% kelleher2019inferring,zhang2023biobank}.
% We address these problems by providing a more flexible formulation
% of an ARG data structure, which we call the ``genome ARG''
% or gARG encoding. We show that it is fully compatible with the
% classical eARG approach, while providing much more flexibility
% in how genetic inheritance can be represented,
% and encompasses the outputs of modern methods.
% Moreover we show that, in contrast with the eARG encoding which
% requires precise estimates of each recombination event,
% the gARG encoding allows us to systematically describe uncertainty
% about the temporal ordering of multiple recombinations.
